Global Distribution of Unintended Pregnancies and Contraceptive Usage

### 2024-04-02 code/analysis-code/statistical-analysis.qmd
Part 4 Updates
Completed bootstrapping and graphed the results. Added LASSO regression and tuned the parameters. Organized all results in the manuscript.

---
title: "Statistical Analysis"
author: "Taylor Glass and Arlyn Santiago"
date: "03/12/24"
output: html_document
---
## Load necessary packages
```{r}
#| message: false
library(dplyr) # for basic syntax
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) # for model building 
library(yardstick) # for finding performance metrics 
library(rsample) #for splitting the data into testing/training
library(knitr) # to create performance metric tables 
library(parsnip) # to create a workflow for cross-validation
```

@@

@@

# collect the metrics from both models
collect_metrics(my_resample)
collect_metrics(my_resample2)
```
The cross-validated results lead to the same conclusion: the multiple linear regression model with 5 predictors performs better. The RMSE value for the single predictor model with `sarc_standard` is 4.822226, which is lower the the first RMSE value of 5.143319. While the RMSE value for the multiple linear regression model is slightly higher than the first time it was calculated, it is still slightly lower at 4.405716. Additionally, the standard error the multiple linear regression model is 0.2768, which is lower than the standard error for the single predictor model of 0.3528. This means there is higher variance in the single predictor model. The R-squared value of 0.8206 for the multiple linear regression model is higher than 0.7861 for the single predictor model, which means the multiple linear regression model accounts for more of the variability in the data. Overall, the multiple linear regression model performs better.

@@

# create a visual representation with 2 sets of predicted and observed values 
ggplot(plotdata, aes(x = observed)) +
  geom_point(aes(y = training_model2, color = "training model"), shape = 3) +
  geom_point(aes(y = test_model2, color = "testing model"), shape = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") + #add the 45 degree line
  xlim(0, 50) + #limit the x-axis values
  ylim(0, 50) + #limit the y-axis values
  labs(x = "Observed Values", y = "Predicted Values", title = "Comparing Training and Testing Data on MLR Model")
```
The training model predictions, shown in teal, hover pretty closely around the expected relationship between the outcome and the 5 predictors, which is depicted as the 45 degree line. The testing model predictions, shown in coral, are much more extreme than the training model predictions because they are farther from the 45 degree line. While there is some overlap between the two sets of predictions, it is likely that our model is overfitting the data.

>>>>

---
title: "Statistical Analysis"
author: "Taylor Glass and Arlyn Santiago"
date: "03/12/24"
output: html_document
---
## Load necessary packages
```{r}
#| message: false
library(dplyr) # for basic syntax
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) # for model building 
library(yardstick) # for finding performance metrics 
library(rsample) #for splitting the data into testing/training
library(knitr) # to create performance metric tables 
library(parsnip) # to create a workflow for cross-validation
library(glmnet) # for LASSO regression
```

@@

We want to have a table comparing the RMSE and MAE values for the three single predictor models of interest along with the full model with all 5 predictors, which will all be compared to the null model. 
```{r}
# create a matrix with RMSE and MAE values
results3 <- as.table(matrix(c(rmse1$.estimate, mae1$.estimate, rmse2$.estimate, mae2$.estimate,
                             rmse3$.estimate, mae3$.estimate, rmse9$.estimate, mae9$.estimate,
                             rmse4$.estimate, mae4$.estimate), nrow = 5, ncol = 2, byrow = TRUE))

# add column names
colnames(results3) <- c("RMSE", "MAE")

# add row names
rownames(results3) <- c("Population Size Using SARCs", "Percent Currently Married", 
                       "Rate of Maternal Deaths due to Abortion", "Full Model with 5 Predictors", "Null Model")

# print the table
kable(results3)

# save performance metrics results table  
table_file14 = here("results", "tables", "resulttable14.rds")
saveRDS(results3, file = table_file14)
```

@@

# collect the metrics from both models
collect_metrics(my_resample)
collect_metrics(my_resample2)
```
The cross-validated results lead to the same conclusion: the multiple linear regression model with 5 predictors performs better. The RMSE value for the single predictor model with `sarc_standard` is 4.822226, which is lower the the first RMSE value of 5.143319. While the RMSE value for the multiple linear regression model is slightly higher than the first time it was calculated, it is still slightly lower at 4.405716. Additionally, the standard error of the RMSE for the multiple linear regression model is 0.2768, which is lower than the standard error of the RMSE for the single predictor model of 0.3528. This means there is higher variance in the single predictor model. The R-squared value of 0.8206 for the multiple linear regression model is higher than 0.7861 for the single predictor model, which means the multiple linear regression model accounts for more of the variability in the data. Overall, the multiple linear regression model performs better.

We want a table displaying these results as well. 
```{r}
# create a matrix with RMSE and MAE values
results4 <- as.table(matrix(c(4.822226, 0.25279125, 0.786067, 0.04292202, 4.405716, 0.27679502, 0.820600, 0.02815933), nrow = 2, ncol = 4, byrow = TRUE))

# add column names
colnames(results4) <- c("RMSE", "Standard Error", "R-Squared", "Standard Error")

# add row names
rownames(results4) <- c("Single Predictor Model", "Full Predictor Model")

# print the table
kable(results4)

# save performance metrics results table  
table_file15 = here("results", "tables", "resulttable15.rds")
saveRDS(results4, file = table_file15)
```

@@

# create a visual representation with 2 sets of predicted and observed values 
figure1 <- ggplot(plotdata, aes(x = observed)) +
              geom_point(aes(y = training_model2, color = "training model"), shape = 3) +
              geom_point(aes(y = test_model2, color = "testing model"), shape = 3) +
              geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") + #add the 45               degree line
              xlim(0, 50) + #limit the x-axis values
              ylim(0, 50) + #limit the y-axis values
              labs(x = "Observed Values", y = "Predicted Values", title = "Comparing Training and Testing Data on MLR Model")
figure1

# save final figure
figure_file = here("results", "figures", "final-figures", "fullmodelpred.png")
ggsave(filename = figure_file, plot=figure1) 
```
The training model predictions, shown in teal, hover pretty closely around the expected relationship between the outcome and the 5 predictors, which is depicted as the 45 degree line. The testing model predictions, shown in coral, are much more extreme than the training model predictions because they are farther from the 45 degree line. While there is some overlap between the two sets of predictions, it is likely that our model is overfitting the data.

To examine the uncertainty in the model further, we will use the bootstrapping method to resample our data 100 times to find confidence intervals for our predictions. 
```{r}
# set the seed for reproducibility
set.seed(2468)

# create 100 bootstraps with the training data
bootstraps <- bootstraps(train_data, times = 100)

# create empty vector to store predictions list 
preds_bs <- vector("list", length = length(bootstraps))

# write a loop to fit the model to each bootstrap and make predictions
for (i in 1:length(bootstraps)) {

  bootstrap_sample <- analysis(bootstraps$splits[[i]])  # isolate the bootstrap sample

  model <- lm(pct_upreg_sarc ~ sarc_standard + region + pct_currentlymarried + rate_matdeaths_upreg + rate_matdeaths_abs, data = bootstrap_sample) # fit the model using the bootstrap sample

  predictions <- predict(model, newdata = train_data) # make predictions with the training data

  preds_bs[[i]] <- predictions # store predictions in the empty vector
}
# create an empty array to store the predictions
num_samples <- length(preds_bs)
num_datapoints <- length(preds_bs[[1]])  
preds_array <- array(NA, dim = c(num_samples, num_datapoints))

# fill the array with predictions from bootstrappping
for (i in 1:num_samples) {
  preds_array[i,] <- unlist(preds_bs[[i]])
}

# find the median and 95% confidence intervals of each prediction
preds <- preds_array %>%
          apply(2, quantile,  c(0.05, 0.5, 0.95)) %>%
          t()

# create a data frame with all the necessary variables
finalbootstrap <- data.frame(
  observed = c(train_data$pct_upreg_sarc), 
  point_estimates = c(trainpred$.pred),
  medians = preds[, 2],
  lower_bound = preds[, 1],
  upper_bound = preds[, 3]
)

# create visualization with all 5 variables 
figure2 <- ggplot(finalbootstrap, aes(x = observed, y = point_estimates)) +
              geom_point(color = "black") +  
              geom_point(aes(y = medians), color = "coral") + 
              geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.1, color = "blue") + 
              geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +  # add 45                 degree line
              xlim(0, 50) + # limit the x-axis values
              ylim(0, 50) + # limit the y-axis values
              labs(x = "Observed Values", y = "Predicted Values", title = "Uncertainty in Predictions with Final Model")
figure2

# save final figure
figure_file2 = here("results", "figures", "final-figures", "bootstrappred.png")
ggsave(filename = figure_file2, plot=figure2) 
```
This figure shows that 95% confidence intervals, shown as dark blue lines, for the predictions that do not fall on the expected 45 degree line do not intersect with the line either. The dark blue lines show that the confidence intervals are narrow, which increases the validity of the conclusion that the model is overfitting the data because there is not much uncertainty in the predictions.

Since we have evidence that our model overfits the data, the last technique we want to use to explore our full model is Lasso regression. This regression analysis shrinks the coefficients towards the mean, which allows it to perform subset selection. Coefficients can reach zero, and that results in that variable being removed from the model. This technique could reveal predictors that should not be included in our final model, which could reduce the risk of overfitting. 
```{r}
# set seed for reproducibility
set.seed(2468)

# LASSO regression workflow
lasso_mod <- linear_reg(penalty = 0.1, mixture = 1, mode = "regression") %>% #set the penalty
                set_engine("glmnet")

lasso_results <- lasso_mod %>% 
            fit(pct_upreg_sarc ~ sarc_standard + region + pct_currentlymarried + rate_matdeaths_upreg +
                  rate_matdeaths_abs, data = mydata)

# print results 
tidy(lasso_results)

# save results from fit into a data frame 
lmtable11 <- tidy(lasso_results)

# save fit results table  
table_file16 = here("results", "tables", "resulttable16.rds")
saveRDS(lmtable11, file = table_file16)
```
As expected, the coefficients are much smaller than the first full model we created without the regularization penalty, but all 5 predictors remain in the model. The `sarc_standard` predictor still has the strongest relationship with the outcome. The percent of women currently married, rate of maternal deaths due to unintended pregnancies, and rate of maternal deaths to due to abortion all have weak negative associations with the outcome. The direction of the relationship between the rate of maternal deaths due to abortions and the outcome flipped from the first full regression model to this model, but the coefficient is so small that this difference is not significant.

It is important to tune the hyperparameters associated with LASSO regression to determine which model fits best. We will use a grid search to ensure that we try all combinations of the parameters in our grid. We will complete the tuning process with a 5-fold cross-validation repeated 5 times. 
```{r}
# set seed for reproducibility
set.seed(2468)

# repeat lasso model workflow
lasso_mod2 <- linear_reg(penalty = tune(), mixture = 1, mode = "regression") %>% # allow for tuning
                set_engine("glmnet")

# change variable type of region for tune_grid()
mydata$region <- as.numeric(mydata$region)

# define the recipe
recipe <- recipe(pct_upreg_sarc ~ sarc_standard + region + pct_currentlymarried + rate_matdeaths_upreg +
                  rate_matdeaths_abs, data = mydata)

# define the workflow
workflow <- workflow() %>%
  add_recipe(recipe) %>%
  add_model(lasso_mod2)

# define a parameter grid with 50 values linearly spaced on a log scale
model_grid <- grid_regular(penalty(range = c(log10(1e-5), log10(1e2))), levels = 50)

# create the real samples cross-validation 
real_samples <- vfold_cv(mydata, v = 5, repeats = 5)

# tune the model 
tuned_model <- workflow %>% 
                tune_grid(resamples = real_samples,
                          grid = model_grid,
                          metrics = metric_set(rmse))

# observe the tuning process
figure3 <- tuned_model %>% 
             autoplot()
figure3

# save the final figure 
figurefile3 <- here("results", "figures", "final-figures", "lassoparameters.png")
ggsave(filename = figurefile3, plot = figure3)
```
As the regularization penalty increases past 1, we see that the RMSE value increases exponentially. The lowest RMSE for the model is slightly greater than 5 when the regularization penalty is less than 1. When the regularization penalty increases to 10, the RMSE increases to almost 11. These observations are expected because the RMSE can only increase in LASSO regression because the penalty parameter is added to the SSR term to determine the value of the cost function. It is expected that the model will perform worse according to the RMSE scale as the penalty parameter increases. This plot shows that penalty parameters less than 1 are preferable for model performance. Since we originally chose 0.1 as our penalty parameter, we found one of the best performing models with the LASSO regression.

### 2024-04-02 products/manuscript/Manuscript.qmd
Part 4 Updates
Completed bootstrapping and graphed the results. Added LASSO regression and tuned the parameters. Organized all results in the manuscript.

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). is an example of some - completely random/unrelated - schematic that was generated with Biorender.
We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
# knitr::include_graphics(here("products","assets","antigen-recognition.png")) #commented out for the sake of knitting the file at this stage in the project 
```

@@

The second model we created used the region predictor, which revealed that the Latin America/the Caribbean region had the strongest relationship with the outcome variable followed closely by the Asia region. The Oceania region does not appear to have a strong effect on the outcome variable with the Europe region only showing a slightly larger association. These associations can be seen in @statistical-analysis2.
```{r}
#| label: statistical-analysis2
#| fig-cap: "Regional Predictor"
#| echo: FALSE
readRDS(here("results","tables", "resulttable2.rds")) 
```
The Tukey HSD test for the region predictor, shown in @statistical-analysis3, revealed the largest differences in the outcome between the Oceania region and Latin America/the Caribbean, Europe, and Asia. The differences between Oceania and Europe and Latin America/the Caribbean are the largest. Oceania is clearly experiencing the lowest percentage of unintended pregnancies among short acting reversible method users. Two more significant differences in the outcome exist between Latin America/the Caribbean and Europe with Africa, which shows that Africa is experiencing higher levels of the outcome. 
```{r}
#| label: statistical-analysis3
#| fig-cap: "Tukey HSD Test Results"
#| echo: FALSE
readRDS(here("results","tables", "resulttable13.rds")) 
```

We created one model with all 5 predictors using multiple linear regression, shown in @statistcal-analysis4. This model shows the strongest predictor is the number of women using short acting reversible contraception. The next strongest predictor is the European region followed by the Latin America/the Caribbean region and the Asia region, which is a shift from the model with the singular region predictor. The Oceania region now has a negative association with the outcome. The currently married predictor is associated with a weak decrease in the outcome. The coefficient estimates for the rate of maternal death due to unintended pregnancies and abortion both show a very weak positive association with the outcome. 
```{r}
#| label: statistical-analysis4
#| fig-cap: "Multiple Linear Regression with 5 Predictors"
#| echo: FALSE
readRDS(here("results","tables", "resulttable8.rds")) 
```
The full statistical analysis can be found in the statistical-analysis.qmd file in the code > analysis-code folder.

@@

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
# resulttable2 = readRDS(here("results","tables-files","resulttable2.rds"))
# knitr::kable(resulttable2) #commented out for the sake of knitting our project at this stage
```

>>>>

@@

The second model we created used the region predictor, which revealed that the Latin America/the Caribbean region had the strongest relationship with the outcome variable followed closely by the Asia region. The Oceania region does not appear to have a strong effect on the outcome variable with the Europe region only showing a slightly larger association. These associations can be seen in @statistical-analysis2.
```{r}
#| label: statistical-analysis2
#| fig-cap: "Regional Predictor"
#| echo: FALSE
readRDS(here("results","tables", "resulttable2.rds")) 
```
Details about the 9 different models we created can be found in the statistical-analysis.qmd file in the code > analysis-code folder along with an exploration of the correlation between predictors.

@@

We wanted to analyze the differences in the outcome between regions further, so we completed an ANOVA test and subsequet Tukey HSD test. The Tukey HSD test for the region predictor, shown in @statistical-analysis3, revealed the largest differences in the outcome between the Oceania region and Latin America/the Caribbean, Europe, and Asia. The differences between Oceania and Europe and Latin America/the Caribbean are the largest. Oceania is clearly experiencing the lowest percentage of unintended pregnancies among short acting reversible method users. Two more significant differences in the outcome exist between Latin America/the Caribbean and Europe with Africa, which shows that Africa is experiencing higher levels of the outcome. 
```{r}
#| label: statistical-analysis3
#| fig-cap: "Tukey HSD Test Results"
#| echo: FALSE
readRDS(here("results","tables", "resulttable13.rds")) 
```
We created one model with all 5 predictors using multiple linear regression, shown in @statistcal-analysis4. This model shows the strongest predictor is the number of women using short acting reversible contraception. The next strongest predictor is the European region followed by the Latin America/the Caribbean region and the Asia region, which is a shift from the model with the singular region predictor. The Oceania region now has a negative association with the outcome. The currently married predictor is associated with a weak decrease in the outcome. The coefficient estimates for the rate of maternal death due to unintended pregnancies and abortion both show a very weak positive association with the outcome. 
```{r}
#| label: statistical-analysis4
#| fig-cap: "Multiple Linear Regression with 5 Predictors"
#| echo: FALSE
readRDS(here("results","tables", "resulttable8.rds")) 
```
To determine which model performed the best, we started by examining the RMSE and MAE performance metrics. We found that the model with the standardized total count of women using the short acting reversible contraception method performed the best in a single predictor model compared to the percent of currently married women and the rate of maternal mortality due to abortion. @statistical-analysis5 shows the RMSE and MAE values for each of these three models compared to the null model with no predictors. The model with the population size using SARCs is the only single predictor model that performs better than the null. The full model with all 5 predictors peforms the best with the lowest RMSE and MAE values. 
```{r}
#| label: statistical-analysis5
#| fig-cap: "Performance Metric Comparisons"
#| echo: FALSE
readRDS(here("results","tables", "resulttable14.rds")) 
```
We used cross-validation of the two best models to determine how well the model performs on data it has not seen yet by splitting the data into testing and training pieces. The cross-validated results, shown in @statistical-analysis6 lead to the same conclusion: the multiple linear regression model with 5 predictors performs better. The RMSE value for the single predictor model with `sarc_standard` is 4.822226, which is lower the the first RMSE value of 5.143319. While the RMSE value for the multiple linear regression model is slightly higher than the first time it was calculated, it is still lower than the single predictor model. Additionally, the standard error of the RMSE for the multiple linear regression model is 0.2768, which is lower than the standard error of the RMSE for the single predictor model of 0.3528. This means there is higher variance in the single predictor model. The R-squared value of 0.8206 for the multiple linear regression model is higher than 0.7861 for the single predictor model, which means the multiple linear regression model accounts for more of the variability in the data. Overall, the multiple linear regression model performs better.
```{r}
#| label: statistical-analysis6
#| fig-cap: "Cross Validation Performance Metric Comparisons"
#| echo: FALSE
readRDS(here("results","tables", "resulttable15.rds")) 
```

Since we determined that the full model with all 5 predictors performs best, we graphed the predicted versus observed values using this model with both the testing and training data to further examine model performance. As seen in @statistical-analysis7, the training model predictions, shown in teal, hover pretty closely around the expected relationship between the outcome and the 5 predictors, which is depicted as the 45 degree line. The testing model predictions, shown in coral, are much more extreme than the training model predictions because they are farther from the 45 degree line. While there is some overlap between the two sets of predictions, it is likely that our model is overfitting the data. 
```{r}
#| label: statistical-analysis7
#| fig-cap: "Evaluating Final Model Performance on Test Data"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "fullmodelpred.png"))
```
We went one step further to examine the uncertainty in the full model predictions using the bootstrap method with 100 resamples of the training data. The figure shown in @statistical-analysis8 displays the 95% confidence intervals, shown as dark blue lines, for the predictions made by the model. The confidence intervals are narrow, which shows there is minimal uncertainty in the model prediction. For the predictions that do not fall on the expected 45 degree line, the confidence intervals do not intersect with the line either. These findings increase the validity of the conclusion that the model is overfitting the data because there is not much uncertainty in the predictions. 
```{r}
#| label: statistical-analysis8
#| fig-cap: "Evaluation of Uncertainty in Full Model Predictions"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "bootstrappred.png"))
```

Since we have evidence that our model overfits the data, the last technique we want to use to explore our full model is Lasso regression. This technique could reveal predictors that should not be included in our final model, which could reduce the risk of overfitting. As seen in @statistical-analysis9, the LASSO regression coefficients are much smaller than the first full model we created without the regularization penalty, but all 5 predictors remain in the model. The `sarc_standard` predictor still has the strongest relationship with the outcome. The percent of women currently married, rate of maternal deaths due to unintended pregnancies, and rate of maternal deaths to due to abortion all have weak negative associations with the outcome. This process confirms that our final model should contain all 5 predictors. 
```{r}
#| label: statistical-analysis9
#| fig-cap: "LASSO Regression Results"
#| echo: FALSE
readRDS(here("results","tables", "resulttable16.rds")) 
```
It is important to tune the hyperparameters associated with LASSO regression to determine the model that fits best, which is shown in @statistical-analysis10. As the regularization penalty increases past 1, we see that the RMSE value increases exponentially. The lowest RMSE for the model is slightly greater than 5 when the regularization penalty is less than 1. When the regularization penalty increases to 10, the RMSE increases to almost 11. Since we originally chose 0.1 as our penalty parameter, we found one of the best performing models with the LASSO regression based on an RMSE value slightly greater than 5, which i comparable to the RMSE found in the first regression model with all 5 predictors. 
```{r}
#| label: statistical-analysis10
#| fig-cap: "Parameter Tuning for LASSO Regression"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "lassoparameters.png"))
```
